Perihelion advance of Mercury

### 2020-10-07 perihelion_advance_of_mercury.Rmd
add OLS analysis to Newtonian case

```{r,label=calculateprecession}
prec <- lm(phi~tee)$coefficients[2]
prec
```

@@

closely agreeing with Einstein's original prediction of 43''.  A more
careful analysis would account for the nonzero mass of Mercury.

By way of a consistency check, we can plot a similar diagram but with
the classical approximation:

```{r,precessionwithclassicaldynamics,cache=TRUE}
out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7),GR=FALSE)
# trajectory() takes ~ 5 mins to run
jj1 <- c(NA,out$rdot)
jj2 <- c(out$rdot,NA)
change <- (jj1>0) & (jj2<0)
tee    <- out$time[change]
phi <- out$phi[change]
phi <- (phi) %% (2*pi)
plot(tee,phi-pi,main='orbital angle at of Mercury at perihelion, classical case',xlab='time (seconds)',ylab='angle of perihelion')
```

In the above plot, note how there is no trend, consistent with the
Newtonian case of zero perihelion advance.

@@

Compare with the value of about 43.37 obtained from numerical
integration, showing close agreement.

>>>>

```{r,label=calculateprecession}
fit <- lm(phi~tee)
summary(fit)
prec <- lm(phi~tee)$coefficients[2]
prec
```

@@

The estimate of closely agreeing with Einstein's original prediction
of 43'' (note that the standard error in `fit` of about
`r round(summary(fit)$coefficients[2,2] * prod(conversion),2)`'' is a
severe overestimate as this ignores the serial correlation in the
residuals); a more careful analysis would account for the nonzero mass
of Mercury.  By way of a consistency check, we can plot a similar
diagram but with the classical approximation:

```{r,precessionwithclassicaldynamics,cache=TRUE}
out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7),GR=FALSE)
# trajectory() takes ~ 5 mins to run
jj1 <- c(NA,out$rdot)
jj2 <- c(out$rdot,NA)
change <- (jj1>0) & (jj2<0)
tee    <- out$time[change]
phi <- out$phi[change]
phi <- (phi) %% (2*pi)
newtfit <- summary(lm(phi~tee))
plot(tee,phi-pi,main='orbital angle at of Mercury at perihelion, classical case',xlab='time (seconds)',ylab='angle of perihelion')
```

In the above plot, note how there is no trend [the OLS estimate of the
slope is $`r round(prod(conversion)*newtfit$coefficients[2,1],2)` \pm
`r round(prod(conversion)*newtfit$coefficients[2,2],2)`$, again in
units of arcseconds per century], consistent with the Newtonian case
of zero perihelion advance.

@@

Compare with the value of about `r round(prec*prod(conversion),2)`
obtained above from numerical integration, showing close agreement.
